---
title       : Spectral Leakage 
subtitle    : Demonstration with Shiny App (JHU Coursera)
author      : Mike Jaffe
job         : Cardiorespiratory Consulting, LLC
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Spectral Leakage

**Fourier Analysis** 
- Waveform approximated by sums of sinusoids
- Important tool with time series analysis 
- FFT digital implementation (fft() in R)
- Assumes sequence repeats (i.e. is periodic) - often leads to discontinuity
- Windowing used to handle the resulting "spectral leakage"

<img src="C:/Users/mjaffe/Documents/CourseraandOpenEDx/_DataScience(Hopkins)/DevelopingDataProducts/spectral/assets/fig/discont.jpg" alt="Maxim"/>
(Adapted from Maxim website)

See http://mbjengineering.shinyapps.io/runApp/ to explore these concepts further. 

--- .class #id 

## Ideal Situation (pure sinusoid with no waveform discontinuity)

Over30 db difference between fundamental frequency and surrounding frequencies
 
```{r, echo=FALSE, fig.width = 12, fig.height = 6}
                offsetdeg    = 0
                samplingrate = 128
                leninsec     = 16
                nbrcycles    = 16
                windowing    = FALSE
                HFCycle      = FALSE
                
                nbrpts = samplingrate*leninsec
                offset = offsetdeg*(2*pi)/360
                
                indices = nbrcycles*c(0:(nbrpts-1))*(2*pi)/(nbrpts)
                waveform = sin(indices+offset)
                
                
                # windowing
                if (windowing) {
                        # Hanning
                        windices = c(0:(nbrpts-1))*((2*pi)/(nbrpts-1))
                        window= 0.5*(1-cos(windices))
                        waveform = window * waveform  
                }
                
                par( mfrow = c( 1, 2 ) )
                timeindices = c(0:(nbrpts-1))/samplingrate
                plot(timeindices,waveform, type='l',xlab="Time(seconds)", col = 'darkgray')

# Fourier Transform
                finput = waveform - mean(waveform)
                fftout = fft(finput)
                realpart = Re(fftout)
                imagpart = Im(fftout)
                magn = sqrt (realpart*realpart+imagpart*imagpart)
                
                # symmetric - toss half and only plot region around peak
                ptstoplot = (nbrpts/2)
                submagn    = magn[1:ptstoplot]
                maxval = max(submagn)
                maxpos = which.max(submagn)
                maxfreq= samplingrate/(maxpos-1)
                if ((maxpos*2) < ptstoplot) {
                        ptstoplot =maxpos*2
                        submagn    = submagn[1:ptstoplot]
                }

                subindices = (0:(ptstoplot-1))*(samplingrate/nbrpts)
                logsubmagn = log(submagn)
                plot(subindices,logsubmagn,ylab="log(Magnitude)",xlab="Frequency(Hz)",type='p')
                
``` 

--- .class #id 

## Discontinuity - pure sinusoid with discontinuity

The fundamental frequency difficult to visualize.

```{r, echo=FALSE, fig.width = 12, fig.height = 6}

                samplingrate = 128
                leninsec     = 16
                nbrcycles    = 16
                HFCycle      = TRUE
                if (HFCycle) nbrcycles = nbrcycles + 0.5 
                indices = nbrcycles*c(0:(nbrpts-1))*(2*pi)/(nbrpts)
                waveform = sin(indices+offset)
                
                
                # windowing
                if (windowing) {
                        # Hanning
                        windices = c(0:(nbrpts-1))*((2*pi)/(nbrpts-1))
                        window= 0.5*(1-cos(windices))
                        waveform = window * waveform  
                }
                
                par( mfrow = c( 1, 2 ) )
                timeindices = c(0:(nbrpts-1))/samplingrate
                plot(timeindices,waveform, type='l',xlab="Time(seconds)", col = 'darkgray')

# Fourier Transform
                finput = waveform - mean(waveform)
                fftout = fft(finput)
                realpart = Re(fftout)
                imagpart = Im(fftout)
                magn = sqrt (realpart*realpart+imagpart*imagpart)
                
                # symmetric - toss half and only plot region around peak
                ptstoplot = (nbrpts/2)
                submagn    = magn[1:ptstoplot]
                maxval = max(submagn)
                maxpos = which.max(submagn)
                maxfreq= samplingrate/(maxpos-1)
                if ((maxpos*2) < ptstoplot) {
                        ptstoplot =maxpos*2
                        submagn    = submagn[1:ptstoplot]
                }

                subindices = (0:(ptstoplot-1))*(samplingrate/nbrpts)
                logsubmagn = log(submagn)
                plot(subindices,logsubmagn,ylab="log(Magnitude)",xlab="Frequency(Hz)",type='p')
                
``` 

--- .class #id 

## Windowing - pure sinusiod with discontinuity and Hanning Window Applied

The fundamental is visible with approx. a 10 db decrease in surrounding frequencies.

```{r, echo=FALSE, fig.width = 12, fig.height = 6}

                samplingrate = 128
                leninsec     = 16
                nbrcycles    = 16         
                windowing    = TRUE
                
                waveform = sin(indices+offset)
                
                
                # windowing
                if (windowing) {
                        # Hanning
                        windices = c(0:(nbrpts-1))*((2*pi)/(nbrpts-1))
                        window= 0.5*(1-cos(windices))
                        waveform = window * waveform  
                }
                
                par( mfrow = c( 1, 2 ) )
                timeindices = c(0:(nbrpts-1))/samplingrate
                plot(timeindices,waveform, type='l',xlab="Time(seconds)", col = 'darkgray')

# Fourier Transform
                finput = waveform - mean(waveform)
                fftout = fft(finput)
                realpart = Re(fftout)
                imagpart = Im(fftout)
                magn = sqrt (realpart*realpart+imagpart*imagpart)
                
                # symmetric - toss half and only plot region around peak
                ptstoplot = (nbrpts/2)
                submagn    = magn[1:ptstoplot]
                maxval = max(submagn)
                maxpos = which.max(submagn)
                maxfreq= samplingrate/(maxpos-1)
                if ((maxpos*2) < ptstoplot) {
                        ptstoplot =maxpos*2
                        submagn    = submagn[1:ptstoplot]
                }

                subindices = (0:(ptstoplot-1))*(samplingrate/nbrpts)
                logsubmagn = log(submagn)
                plot(subindices,logsubmagn,ylab="log(Magnitude)",xlab="Frequency(Hz)",type='p')
                
``` 